Deriving an Executable Model of a Data Stream Protocol from its Specification

### 2009-09-24 papers/ModelBasedTestCaseGenerationDataStreamProtocol/paper.tex
ModelBasedTestCaseGenerationDataStreamProtocol paper: added JML specification examples

%
% "The UCD CASL SenseTile System", for IEEE Pervasive Computing
% Joseph R. Kiniry, Vieri del Bianco, Dragan Stosic
% $Id: paper.tex 2795 2007-09-23 19:35:53Z dmz $
%

@@

\usepackage{xspace}
\usepackage{tabularx}
\usepackage{epsfig}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{eucal}
\usepackage{stmaryrd}
\usepackage{float}

@@

@@

\todo other approaches to the problem

@@

\todo different layers need different simulators

\todo describe simulators structure

\subsection{A JML specification example}
\label{subsec:a_jml_specification_example}

\todo find interesting JML specification

\todo describe the example

@@

>>>>

%
% "The UCD CASL SenseTile System", for IEEE Pervasive Computing
% Joseph R. Kiniry, Vieri del Bianco, Dragan Stosic
% $Id: paper.tex 2795 2007-09-23 19:35:53Z dmz 
%

@@

\usepackage{xspace}
\usepackage{tabularx}
\usepackage{epsfig}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{eucal}
\usepackage{stmaryrd}
\usepackage{float}
\usepackage{listings}
\input{jml-listings}
\lstset{language=[JML]Java,xleftmargin=20pt,xrightmargin=20pt}

@@

\newcommand{\lil}[1]{\texttt{\lstinline|#1|}}

@@

\todo search: other approaches to similar problems

@@

Simulators are needed to test in isolation parts of the system; especially the upper 
layers, yet to be developed. 
During the development the simulators were essential, since the board was not available.
Simulators were built concurrently with the specification, the real driver code, and the 
tests.

Theoretically a separate simulator should be needed for each protocol layer; a
separate simulator for each layer would enable us to separeately specify, develop and test
each layer of the protocol.
In our case we evaluated not worth the effort to maintain such a separation in the 
simulators, motivated by the fact that the protocol layers identified are too fine grained 
to be separated into different simulators. 
Two simulators were built.

The two simulators were built reflecting the software abstraction layers defined for the 
driver.
The driver API is splitted in two structurual layers: the general high level interfaces, that 
expose the main functionalities of the board and the main contents of the packets hiding the 
details, and the lower level implementation that parse the data streams in and out the board, 
to translate the higher level intructions and data in properly formed packets.
Hence, we have implemented two simulators: one high level simulator, whicj reimplements the 
high level interfaces providing corresponding high level mechanisms to control the data and 
the behaviour of the simulated board, it has nothing to do with the details of the data 
streams.
One low level simulator which is capable to rebuild the data stream out of the board based on 
the data stream into the board, the in and out data streams are built exaclty as the sensor 
board is expected to be able to parse or generate.

\subsection{JML specification examples}
\label{subsec:a_jml_specification_example}

The JML specifications used to annotate Java code are of varing complexities.
Some of the specifications are simple, focusing on constraints that should hold when 
calling a method (preconditions) and the constraints on the return value (a very basic form of 
postconditions).
In listing \ref{lst:jml_example_simple} a simple JML specification example is shown, the 
method \lil{getTemperature} is declared to be \lil{/*@ pure @*/}, which means that 
it cannot change the state of any object (a postcondition), and the lower and upper bound of 
its return value are defined (another postcondition).

\begin{lstlisting}[
  caption={A simple specification with JML annotation: simple postconditions.},
  label=lst:jml_example_simple,
  float=htbp]
/*@ 
  ensures \result >= -880;
  ensures \result <= 2047;
@*/
/*@ pure @*/ short getTemperature();
\end{lstlisting}

Some of the specifications are rather complex, usually focusing on properties regarding the 
behaviour of a whole object.
In listing \ref{lst:jml_example_invariant} a complex invariant 
(a property that has to be maintained, that is, that is assumed on entry and guaranteed on 
exit of each method of an object)
example is shown.

\sloppy

The invariant is constraining the number of samples for each medium data rate streams: 
the total number of streams is \lil{Frame.ADC_CHANNELS}; the total number of frames is 
\lil{FRAMS}; the constant constraining the number of samples is 
\lil{FRAMES/Frame.ADC_CHANNELS+1}, it means that the samples contained in a frame are
fairly distributed on the channels. 
The valid samples are counted parsing all the frames contained in a packet, selecting only 
the matching valid samples.
A medium data rate stream sample is considered valid when \lil{isADCActive()} method 
returns \lil{true}.

\fussy

\begin{lstlisting}[
  caption={A specification with JML annotation: a complex object invariant which constraints 
    the number of samples for each medium data rate stream channel.},
  label=lst:jml_example_invariant,
  float=htbp]
/*@ 
  invariant (
    \forall int channel;
    0 <= channel && channel < Frame.ADC_CHANNELS; 
    (
      \num_of int i;
      0<= i && i<(FRAMES-1);
      (
        getFrame(i).isADCActive() &&
        (getFrame(i).getADCChannel() == channel)
      )
    ) <= (FRAMES / Frame.ADC_CHANNELS + 1)
  );
@*/
\end{lstlisting}

@@

\todo reason about the usual structure oracle-driver-test-stub(s). a simulator is sometimes a 
stub, sometimes a driver, sometimes an oracle...

\todo insert references to testing embedded systems with simulators and/or with driver/stubs
